Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-10-16 010-background.Rmd
migrating ESCO 2020 talk into sections, revising everything

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. For continuous marginals, the measure works well due to zero probability of ties. For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

@@

@@

>>>>

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. For continuous marginals, the measure works well due to zero probability of ties.

<!-- 
For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

@@

-->

@@

## Other `R` packages for simulating random vectors

There are a few R packages for multivariate simulation.

* `copula` Highly flexible copula specification [@Yan2007]
* `nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001] 
* `Genord` Simulated correlated discrete variables [@BF17]
* `mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]

### 2020-10-16 020-simulation-algorithm.Rmd
migrating ESCO 2020 talk into sections, revising everything

@@

The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss). We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs]. To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:

@@

For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix. Let the unscaled Spearman correlation coefficients be $\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)$ for two marginal distributions and divide the target correlation by the product in the denominator of Equation \@ref(eq:spearmanRescaled). Let these adjustment factors be denoted as $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ and specifically rescale the target Spearman correlation matrix by

@@

In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme. In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$. For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times. Others have used efficient Monte Carlo integration schemes (see @Chen2001), but scale poorly to large dimension in reasonable times (property **S2**). Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.

@@

## Algorithm notes

- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate whenever possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

test workflow again

>>>>

```{r previewChapter020, echo = FALSE, eval = FALSE}
bookdown::preview_chapter('020-simulation-algorithm.Rmd')
```

@@

The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss) to provide a high-performance NORTA (NORmal To Anything) algorithm.

To simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$,
there is the well-known NORTA algorithm [@Cario1997]. It follows like this:

1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ to corresponds with the
   specified output $\Sigma_{\bf Y}$ [@Chen2001; @Xia17]
3. Produce the Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.

We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs]. 
To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:

@@

<!-- 
For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix. Let the unscaled Spearman correlation coefficients be $\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)$ for two marginal distributions and divide the target correlation by the product in the denominator of Equation \@ref(eq:spearmanRescaled). Let these adjustment factors be denoted as $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ and specifically rescale the target Spearman correlation matrix by

@@

your comment -->

In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme. In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$ (see Xia17 and MB13). 
For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times. 
Others have used efficient Monte Carlo integration schemes (see @Chen2001), but scale poorly to large dimension in reasonable times (property **S2**). 
Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- 
that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
We'll study the robustness of our method to this limitation later in the application section (Section BLAH).

@@

One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation (if done pairwise) or through the algorithm's 
bivariate transformations.

### 2020-10-16 040-monte-carlo-evaluation.Rmd
migrating ESCO 2020 talk into sections, revising everything

# Monte Carlo evaluation and Comparsions to other software

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance in comparison to other existing implementations. We focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters. The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example. We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions. We conclude the simulation studies with $10^5$ replicated data sets with sample sizes and distributional characteristics corresponding the estimates from the motivating example data.

The CPU-based computation times we report were from runs using eight dual-threaded 3.9GHz
Xeon Gold processors on a linux workstation, allowing for parallel instructions
over 16 cores. We chose 15 cores for our experiments.

## Other multivariate simulation software

Describe the competing software and algorithms briefly.

- `copula`
- `Genord`
- `Multiord`
- `nortaRA` only does Pearson matching.

@@

### Continuous example: Bivariate Exponential

Both Spearman's and Kendall's measures of dependency are invariant under
strictly monotone transformations (ref). This in turn provides exact simulation
of continuous marginals $F_i$ $i=1,\ldots,d$ since their corresponding quantile
function are increasing monotonically over their support. The results below
provide computations times for $N=10^5$ under increasingly large dimension $d$.
For simplicity we assume multivariate Normal with varying parameters and to
additionally demonstrate the utility of Kendall's $\tau$ even in the Gaussian
setting (as opposed to the nature choice of Pearson's correlation). To layer on complexity, we now assume a multivariate gamma distribution with fixed marginal parameters.

Compare speed and accuracy.

```{r tabGamma, echo = TRUE}
## computation times normal

```
### Discrete example: Bivariate Geometric

Compare speed and accuracy.

## Simulation II: Ultra HD multivariate simulation with identical margins and exchangeable rank-based correlation.

### UHD Continuous example: Multivariate Gamma

Compare speed and accuracy.

### UHD Discrete example: Multivariate Negative Binomial

Compare speed and accuracy.

## Simulation III: Ultra HD multivariate simulatino with heterogenous margins and exchangeable rank-based correlation.

### UHD Heteregeous continuous example: Multivariate Gamma

Compare speed and accuracy.

### UHD Heteregeousdiscrete example: Multivariate Negative Binomial

Compare speed and accuracy.

>>>>

# Monte Carlo evaluations

```{r LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
## knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
## reticulate::use_condaenv("bigsimr")
library(bigsimr)
library(tidyverse)
## CORES <- as.integer( parallel::detectCores() - 1 )
set.seed(10162020)
```

```{r previewChapter040, echo = FALSE, eval = FALSE}
bookdown::preview_chapter('040-monte-carlo-evaluation.Rmd')
```

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance in comparison to other existing implementations. 
We focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters. 
The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example. 
We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions to test the robustness of our algorithm since exact matching is not guaranteed.

@@

### Bivariate Normal

* Let's simulate a bivariate normal and check our correlation matching
  performance as N increases.
* Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ )
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

```{r biNormal, echo = FALSE, eval = TRUE}
mom_norm <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(mean = m, sd = s)
}

mu <- 10
sigma <- 1
margins <- alist(
    qnorm(mean = mu, sd = sigma),
    qnorm(mean = mu, sd = sigma)
)

type <- c("pearson", "spearman", "kendall")
Cores <- as.integer( parallel::detectCores() - 1 )
n <- c(1e3, 1e4, 1e5)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
grid_steps <- 100
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()

for (i in 1:nrow(sim_pars)) {
    type <- sim_pars$type[i]
    cores <- sim_pars$cores[i]
    n <- sim_pars$n[i]
    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
    tmp_bounds <- cor_bounds(margins, type = type)
    cor_lo <- tmp_bounds$lower[1,2] + eps
    cor_hi <- tmp_bounds$upper[1,2] - eps
    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
    ## rho = cor_seq[1]
    for (rho in cor_seq) {
        Rho <- matrix(rho, 2, 2)
        diag(Rho) <- 1.0
        ## Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
        time_data <- system.time({
            ## x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
            ## use rvec to test our software
            x <- rvec(n = n,
                      rho = Rho,
                      margins = margins,
                      type = type,
                      cores = cores)
        })
        ## Save the sims in case
        id <- paste0(
            "d", 2,
            "-N", n,
            "-c", cores,
            "-r", rho,
            "-Cor", type,
            "-adj", as.character(adjustForDiscrete),
            "-dev", "1CORE",
            "-lib", "bigsimr"
        )
        ## save sims in their own subdirectory
        dir.create("./results/sims_norm", showWarnings = FALSE)
        saveRDS(x, file = paste0("./results/sims_norm/", id, ".rds"))

        ## Estimate statistics
        Rho_hat <- cor_fast(x, method = type)
        rho_hat <- Rho_hat[1, 2]
        norm_args_hat <- mom_norm(x[,1])
        mu_hat <- norm_args_hat$mean
        sigma_hat <- norm_args_hat$sd

        ## Save the results
        res <- rbind(res, data.frame(
                              method = "bigsimr",
                              device = "CPU",
                              type = type,
                              cores = cores,
                              margins = "norm",
                              adjustForDiscrete = adjustForDiscrete,
                              d = 2,
                              N = n,
                              rho = rho,
                              rho_hat = rho_hat,
                              mean = mu,
                              sd = sigma,
                              mean_hat = mu_hat,
                              sd_hat = sigma_hat,
                              sim_time = unname(time_data["elapsed"])
                          ))
    }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "./results/norm_sims.rds")
```

----

```{r biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
dat <- readRDS("./results/norm_sims.rds")
dat %>%
    filter(cores == 1) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()

## ggsave('fig/plot-biNormPlot.pdf')
```

## Bivariate Gamma

* Similarly, let's check the performance for a non-symmetric continuous distribution: a
  standard (rate =1) bivariate gamma.
* Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

```{r biGamma, echo = FALSE, eval = FALSE}
mom_gamma <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(shape = m^2 / s^2, rate = m / s^2)
}

shape <- 10
rate <- 1
margins <- list(
  list("gamma", shape = shape, rate = rate),
  list("gamma", shape = shape, rate = rate)
)

type <- c("pearson", "spearman", "kendall")
cores <- c(1L)
n <- c(1e3, 1e4, 1e5)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
grid_steps <- 100
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()
for (i in 1:nrow(sim_pars)) {

  type <- sim_pars$type[i]
  cores <- sim_pars$cores[i]
  n <- sim_pars$n[i]
  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]

  tmp_bounds <- computeCorBounds(margins, type = type)
  cor_lo <- tmp_bounds$lower[1,2] + eps
  cor_hi <- tmp_bounds$upper[1,2] - eps
  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

  for (rho in cor_seq) {
    Rho <- matrix(rho, 2, 2)
    diag(Rho) <- 1.0

    time_data <- system.time({
      x <- rvec(n = n,
                rho = Rho,
                params = margins,
                cores = cores,
                type = type,
                adjustForDiscrete = adjustForDiscrete)
    })

    # Save the sims in case
    id <- paste0(
      "d", 2,
      "-N", n,
      "-c", cores,
      "-r", rho,
      "-Cor", type,
      "-adj", as.character(adjustForDiscrete),
      "-dev", "1CORE",
      "-lib", "bigsimr"
    )
    saveRDS(x, file = paste0("sims_gamma/", id, ".rds"))

    # Estimate statistics
    Rho_hat <- fastCor(x, method = type)
    rho_hat <- Rho_hat[1, 2]
    gamma_args_hat <- mom_gamma(x[,1])
    shape_hat <- gamma_args_hat$shape
    rate_hat <- gamma_args_hat$rate

    # Save the results
    res <- rbind(res, data.frame(
      method = "bigsimr",
      device = "1CORE",
      type = type,
      cores = cores,
      margins = "gamma",
      adjustForDiscrete = adjustForDiscrete,
      d = 2,
      N = n,
      rho = rho,
      rho_hat = rho_hat,
      shape = shape,
      rate = rate,
      shape_hat = shape_hat,
      rate_hat = rate_hat,
      sim_time = unname(time_data["elapsed"])
    ))
  }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "gamma_sims_ags.rds")
```

---

```{r biGammaPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
dat <- readRDS("gamma_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'pearson') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r biGammaPlot2, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers the correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
dat <- readRDS("gamma_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'spearman') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r biGammaPlot3, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers the correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
dat <- readRDS("gamma_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'kendall') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

## Bivariate Negative Binomial

* Let's check the performance for a discrete distribution: a bivariate negative binomial
* Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ )
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

```{r biNegBin, echo = FALSE, eval = FALSE}
mom_nbinom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(size = m^2 / (s^2 - m), prob = m / s^2)
}

size <- 4
prob <- 0.5
margins <- list(
  list("nbinom", size = size, prob = prob),
  list("nbinom", size = size, prob = prob)
)

type <- c("pearson", "spearman", "kendall")
cores <- c(1)
n <- c(1e3, 1e4, 1e5)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
grid_steps <- 100
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()
for (i in 1:nrow(sim_pars)) {

  type <- sim_pars$type[i]
  cores <- sim_pars$cores[i]
  n <- sim_pars$n[i]
  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]

  tmp_bounds <- computeCorBounds(margins, type = type)
  cor_lo <- tmp_bounds$lower[1,2] + eps
  cor_hi <- tmp_bounds$upper[1,2] - eps
  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

  for (rho in cor_seq) {
    Rho <- matrix(rho, 2, 2)
    diag(Rho) <- 1.0

    time_data <- system.time({
      x <- rvec(n = n,
                rho = Rho,
                params = margins,
                cores = cores,
                type = type,
                adjustForDiscrete = adjustForDiscrete)
    })

    # Save the sims in case
    id <- paste0(
      "d", 2,
      "-N", n,
      "-c", cores,
      "-r", rho,
      "-Cor", type,
      "-adj", as.character(adjustForDiscrete),
      "-dev", "1CORE",
      "-lib", "bigsimr"
    )
    saveRDS(x, file = paste0("sims_nb/", id, ".rds"))

    # Estimate statistics
    Rho_hat <- fastCor(x, method = type)
    rho_hat <- Rho_hat[1, 2]
    nbinom_args_hat <- mom_nbinom(x[,1])
    size_hat <- nbinom_args_hat$size
    prob_hat <- nbinom_args_hat$prob

    # Save the results
    res <- rbind(res, data.frame(
      method = "bigsimr",
      device = "1CORE",
      type = type,
      cores = cores,
      margins = "nbinom",
      adjustForDiscrete = adjustForDiscrete,
      d = 2,
      N = n,
      rho = rho,
      rho_hat = rho_hat,
      size = size,
      prob = prob,
      size_hat = size_hat,
      prob_hat = prob_hat,
      sim_time = unname(time_data["elapsed"])
    ))
  }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "nbinom_sims_ags.rds")
```

---

```{r  biNegBinPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("nbinom_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'pearson') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r  biNegBinPlot2, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("nbinom_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'spearman') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r  biNegBinPlot3, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("nbinom_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'kendall') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

### 2020-10-16 050-motivating-application-and-examples.Rmd
migrating ESCO 2020 talk into sections, revising everything

# Some applications

```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
## knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
reticulate::use_condaenv("bigsimr")
library(bigsimr)
library(tidyverse)
CORES <- as.integer( parallel::detectCores() - 1 )
set.seed(10142020)
```

@@

```{r processBRCA, echo=FALSE, eval=FALSE, results='none'}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
dim(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.75
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
print(d <- length(genesToKeep))
## save data as rds
saveRDS(brca, 'data/brca.rds')
```

```{r readBRCA, echo=TRUE, eval=TRUE}
## full workflow simulating RNA-seq data
brca <- readRDS(file = 'data/brca.rds') 
```

@@

```{r estRhoBRCA, echo=TRUE, eval=TRUE}
## 1. Estimate Spearman's correlation on the count data
corType <- 'pearson'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

---

@@

```{r runBRCA, results='none'}
n <- 20000
## d <- 100
## n <- 100
## d <- 5
nb_margins <- make_nbinom_alist(sizes, probs)
## compute theoretical bounds to make sure input is okay
## cor_bounds( nb_margins, type = type )
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType, ensure_PSD = TRUE, cores = CORES)
## sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = "pearson", ensure_PSD = TRUE, cores = 4L)
## pairs(sim_nbinom)
nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
cat('The quadratic loss from the true correlation matrix is ') 
norm(nb_Rho - nb_Rho_hat, type = "2")
```

@@

```{r fig:BRCA, echo=F, eval=T, caption='The T-Poisson strategy produces simulated random vectors from a multivariate negative binomial (NB) that replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes).'}
par(mfrow=c(1,3))
plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
     xlim = c(-1, 1), ylim = c(-1, 1),
     pch = 4, cex = 1, col = rgb(0, 0, 1, 0.05),
     xlab = "Target Correlation", ylab = "Estimated Correlation",
     main = "bigsimr Simulation")
abline(a = 0, b = 1, col="red", lty="dashed")
nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
plot(sizes, nbinom_size_prob_hat[1,],
     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
     xlab = "Target Size", ylab = "Estimated Size",
     main = "Direct NB Simulation")
abline(a=0, b=1, col="red", lty="dashed")
plot(log(probs), log(nbinom_size_prob_hat[2,]),
     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
     main = "Direct NB Simulation")
abline(a=0, b=1, col="red", lty="dashed")
par(mfrow=c(1,1))
```

>>>>

# Motivating application and examples

```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
reticulate::use_condaenv("bigsimr")
library(bigsimr)
library(tidyverse)
CORES <- as.integer( parallel::detectCores() - 1 )
set.seed(10142020)
```

```{r previewChapter050, echo = FALSE, eval = FALSE}
bookdown::preview_chapter('050-some-applications.Rmd')
```

@@

```{r processBRCA, echo=FALSE, eval=TRUE, results='none'}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
dim(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.99
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
print(d <- length(genesToKeep))
## save data as rds
## saveRDS(brca, './data/brca99.rds')
```

```{r readBRCA, echo=FALSE, eval=FALSE}
## full workflow simulating RNA-seq data
brca <- readRDS(file = './data/brca99.rds') 
```

@@

```{r estRhoBRCA, echo=TRUE, eval=TRUE}
## 1. Estimate Spearman's correlation on the count data
## corType <- 'pearson'
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

@@

```{r runBRCA, echo=TRUE, results=TRUE, cache = TRUE}
## Set the number of random vectors
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                                ensure_PSD = TRUE, cores = CORES) )
```

@@

```{r figBRCA, echo=F, eval=T, fig.cap="bigsimr produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes)."}
par(mfrow=c(1,3))
## estimate rho from sims and plot
nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
     xlim = c(-1, 1), ylim = c(-1, 1),
     pch = 4, cex = 1, col = rgb(0, 0, 1, 0.05),
     xlab = "Target Correlation", ylab = "Estimated Correlation",
     main = "Correlation Matching")
abline(a = 0, b = 1, col="red", lty="dashed")
## estimate prob,size from sims and plot
nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
plot(sizes, nbinom_size_prob_hat[1,],
     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
     xlab = "Target Size", ylab = "Estimated Size",
     main = "Marginal NB Size Matching")
abline(a=0, b=1, col="red", lty="dashed")
plot(log(probs), log(nbinom_size_prob_hat[2,]),
     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
     main = "Marginal NB Probability Matching")
abline(a=0, b=1, col="red", lty="dashed")
par(mfrow=c(1,1))
```
